---
title: "ggasym"
author: "Joshua H. Cook"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggasym}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(0)
```

```{r load_libraries, warning=FALSE, message=FALSE}
library(dplyr)
library(magrittr)
library(tibble)
library(ggplot2)
library(ggasym)
```

`ggasym` (pronounced "gg-awesome") is a package that offers a new geom, `geom_asymmat` (for "asymmetric matrix"), for the plotting of two fill or color aesthetics on a symmetric matrix. This is likely to be of most use for multi-way statistical tests of multiple groups. Thus, a matrix can be formed of each group along the x and y-axes and the top-left and bottom-right triangles can be filled for the difference in mean values and p-values respectively.

Below, I begin with a simple example to show how easy `geom_asymmat` is to use for anyone familiar with ggplot2. In another vignette (in progress), I show how the results from a multi-way statistical test can be displayed using `ggasym` and the wrapper function *[insert name of stats test wrapper function here]*.

For information on using ggplot2, start [here](https://ggplot2.tidyverse.org/index.html).

---

## Basic Usage

Here is a basic example. `tib` is a "tibble" (ie. fancy "data.frame") of comparisons between groups "A" through "E". There are two values to be plotted, `val_1` and `val_2`, that hold data on the comparison between `g1` and `g2`. `tib` is first passed to `asymmetrise` to fill in all the missing combinations between `g1` and `g2` such that the symmetric matrix can be built. All added values take the value `NA`. The modified data table is finally passed to `ggplot` and `geom_asymmat` is added on. Here, `asymmetrise` added the rows where `g1` and `g2` are equal, thus will fill the diagonal. I will set these values to `val_3`.

```{r example1}
tib <- tibble(g1 = c("A", "A", "A", "A", "B", "B", "B", "C", "C", "D"),
              g2 = c("B", "C", "D", "E", "C", "D", "E", "D", "E", "E"),
              val_1 = seq(1, 10, 1),
              val_2 = rnorm(10, mean = 0, sd = 3))
tib <- asymmetrise(tib, g1, g2)
tib$val_3 <- runif(nrow(tib))
ggplot(tib, aes(x = g1, y = g2)) +
    geom_asymmat(aes(fill_tl = val_1, fill_br = val_2, fill_diag = val_3)) +
    scale_fill_tl_gradient(low = "lightpink", high = "tomato") +
    scale_fill_br_gradient(low = "lightblue1", high = "dodgerblue") +
    scale_fill_diag_gradient(low = "yellow", high = "orange3")
```

The new aesthetics `fill_tl`, `fill_br`, and `fill_diag` behave just like the normal `fill`, except that they correspond to the top-left ("tl") and bottom-right ("br") triangles of the matrix, respectively. This package also includes analogous functions for scaling the fill colors such as `scale_fill_tl_gradient2` and `scale_fill_br_gradientn` that operate just as expected when using ggplot2.

```{r example2}
ggplot(tib, aes(x = g1, y = g2)) +
    geom_asymmat(aes(fill_tl = val_1, fill_br = val_2, fill_diag = val_3)) +
    scale_fill_tl_gradient(low = "lightpink", high = "tomato") +
    scale_fill_br_gradient2(low = "orange", mid = "white", high = "dodgerblue") +
    scale_fill_diag_gradientn(colors = rainbow(25))
```

Of note, with three colorbars, it may be useful to control their position. This can be done just like normal in ggplot by passing the correct values to the `guide` paramter in `scale_fill_*_gradient` ([original documentation](https://ggplot2.tidyverse.org/reference/guide_colourbar.html)). Below are a few of the options where I put the pars horizontal, adjust the ordering, and put the title above each.

```{r}
ggplot(tib, aes(x = g1, y = g2)) +
    geom_asymmat(aes(fill_tl = val_1, fill_br = val_2, fill_diag = val_3)) +
    scale_fill_tl_gradient(low = "lightpink", high = "tomato",
                           guide = guide_colourbar(direction = "horizontal",
                                                   order = 1,
                                                   title.position = "top")) +
    scale_fill_br_gradient2(low = "orange", mid = "white", high = "dodgerblue",
                           guide = guide_colourbar(direction = "horizontal",
                                                   order = 3,
                                                   title.position = "top")) +
    scale_fill_diag_gradientn(colors = rainbow(25),
                           guide = guide_colourbar(direction = "horizontal",
                                                   order = 2,
                                                   title.position = "top"))
```


Since the new geom is a normal ggplot2 object, it can be introduced into a standard ggplot2 workflow. Note that the labels can be adjusted like normal using the `labs` function and using the `fill_tl` and `fill_br` arguments.

```{r example3}
ggplot(tib, aes(x = g1, y = g2)) +
    geom_asymmat(aes(fill_tl = log(val_1),
                     fill_br = val_2,
                     fill_diag = val_3)) +
    scale_fill_tl_gradient(low = "lightpink", high = "tomato",
                           guide = guide_colourbar(direction = "horizontal",
                                                   order = 1,
                                                   title.position = "top")) +
    scale_fill_br_gradient(low = "lightblue1", high = "dodgerblue",
                           guide = guide_colourbar(direction = "horizontal",
                                                   order = 3,
                                                   title.position = "top")) +
    scale_fill_diag_gradient(low = "white", high = "black",
                           guide = guide_colourbar(direction = "horizontal",
                                                   order = 2,
                                                   title.position = "top")) +
    labs(fill_tl = "top-left fill",
         fill_br = "bottom-right fill",
         fill_diag = "diagonal fill",
         title = "Example of ggasym") +
    theme_bw() +
    theme(axis.title = element_blank(),
          plot.title = element_text(hjust = 0.5),
          panel.background = element_rect(fill = "grey70"),
          panel.grid = element_blank()) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0))
```


### Facetting

If you have multiple categories, facetting works as expected. The only difference is in the preparation of the data table: you must `group_by` the value(s) you will facet by before passing to `asymmetrise`. This is shown below.

```{r facetting_setup}
tib <- tibble(g1 = rep(c("A", "A", "B"), 2),
              g2 = rep(c("B", "C", "C"), 2),
              val_1 = seq(1, 6),
              val_2 = rnorm(6),
              grps = c(1, 1, 1, 2, 2, 2))
tib
```

Grouping first by `grps`, the tibble is asymmetrized. I then added values to the diagonal.

```{r grouped_asymm}
tib <- tib %>% group_by(grps) %>% asymmetrise(g1, g2) %>% ungroup()
tib <- tib %>% mutate(val_3 = ifelse(g1 == g2, runif(nrow(tib)), NA))
tib[1:10, ]
```


```{r facetting_plot}
ggplot(tib, aes(x = g1, y = g2)) +
    geom_asymmat(aes(fill_tl = log(val_1),
                     fill_br = val_2,
                     fill_diag = val_3)) +
    scale_fill_tl_gradient(low = "lightpink", high = "tomato",
                           guide = guide_colourbar(direction = "horizontal",
                                                   order = 1,
                                                   title.position = "top")) +
    scale_fill_br_gradient(low = "lightblue1", high = "dodgerblue",
                           guide = guide_colourbar(direction = "horizontal",
                                                   order = 3,
                                                   title.position = "top")) +
    scale_fill_diag_gradient(low = "white", high = "black",
                           guide = guide_colourbar(direction = "horizontal",
                                                   order = 2,
                                                   title.position = "top")) +
    labs(fill_tl = "top-left fill",
         fill_br = "bottom-right fill",
         fill_diag = "diagonal fill",
         title = "Example of facetting with ggasym") +
    theme_bw() +
    theme(axis.title = element_blank(),
          plot.title = element_text(hjust = 0.5),
          panel.background = element_rect(fill = "grey70"),
          panel.grid = element_blank()) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0)) +
    facet_grid(. ~ grps)
```


