---
title: "ggasym"
author: "Joshua H. Cook"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggasym}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, warning=FALSE, message=FALSE}
library(tibble)
library(ggplot2)
library(ggasym)
```

`ggasym` (pronounced "gg-awesome") is a package that offers a new geom, `geom_asymmat` (for "asymmetric matrix"), for the plotting of two fill or color aesthetics on a symmetric matrix. This is likely to be of most use for multi-way statistical tests of multiple groups. Thus, a matrix can be formed of each group along the x and y-axes and the top-left and bottom-right triangles can be filled for the difference in mean values and p-values respectively.

Below, I begin with a simple example to show how easy `geom_asymmat` is to use for anyone familiar with ggplot2. In another vignette (in progress), I show how the results from a multi-way statistical test can be displayed using `ggasym` and the wrapper function *[insert name of stats test wrapper function here]*.

For information on using ggplot2, start [here](https://ggplot2.tidyverse.org/index.html).

---

## General Use

`tib` is a simple tibble object showing two values, `val_1` and `val_2`, for each comparison of the three groups "A", "B", and "C".

```{r}
tib <- tibble(g1 = c("A", "A", "A", "A", "B", "B", "B", "C", "C", "D"),
              g2 = c("B", "C", "D", "E", "C", "D", "E", "D", "E", "E"),
              val_1 = seq(1, 10, 1),
              val_2 = rnorm(10, mean = 0, sd = 3))
tib[1:5, ]
```

### Scale fill gradient

This data is then plotted with two different fill values, one for the top-left (`tl`) triangle and the other for the bottom-right (`br`) triangle. To demonstrate the complete incorporation of normal ggplot2 conventions, the labels for the color scales can be declared like normal using the `labs` function.

```{r}
ggplot(tib) +
    geom_asymmat(aes(x = g1, y = g2, fill_tl = val_1, fill_br = val_2)) +
    scale_fill_tl_gradient(low = "lightpink", high = "tomato") +
    scale_fill_br_gradient(low = "lightblue1", high = "dodgerblue") +
    labs(fill_tl = "top-left fill", fill_br = "bottom-right fill")
```

### Scale fill gradient with a midpoint

```{r}
ggplot(tib) +
    geom_asymmat(aes(x = g1, y = g2, fill_tl = val_1, fill_br = val_2)) +
    scale_fill_tl_gradient(low = "lightpink", high = "tomato") +
    scale_fill_br_gradient2(low = "gold", high = "dodgerblue", mid = "white") +
    labs(fill_tl =  "top-left fill", fill_br = "bottom-right fill")
```

### Adjusting theme and other elements

Since `geom_asymmat` is just like a normal ggplot geom, it can be integrated into a normal ggplot assembly pipeline. Here is an example where I change the x and y-axes, add a background color, remove the gridlines, and add a title center-justified. I also demonstrate the full integration by passing `log(val_1)` to the  `fill_tl` aesthetic.

```{r}
ggplot(tib) +
    geom_asymmat(aes(x = g1, y = g2, fill_tl = log(val_1), fill_br = val_2)) +
    scale_fill_tl_gradient(low = "lightpink", high = "tomato") +
    scale_fill_br_gradient(low = "lightblue1", high = "dodgerblue") +
    labs(fill_tl = "top-left fill", fill_br = "bottom-right fill",
         title = "Example of ggasym") +
    theme(axis.title = element_blank(),
          plot.title = element_text(hjust = 0.5),
          panel.background = element_rect(fill = "grey50"),
          panel.grid = element_blank()) +
    scale_x_discrete(expand = c(0, 0)) +
    scale_y_discrete(expand = c(0, 0))
```

