---
title: "Statistical Test Plotting with 'ggasym'"
author: "Joshua H. Cook"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical Test Plotting with 'ggasym'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5,
    fig.align = "center"
)
set.seed(0)
```

## Purpose

One of the great uses of 'ggasym' is to plot two values from the results of a multi-way statistical test. Each comparison is a cell, and two values can be used for the fills. Below I give brief examples and plot the differences in mean and the *p*-value on the symmetric matrix.

```{r load_libs, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(tibble)
library(purrr)
library(broom)
library(ggasym)
```

## Data

The data will be modeled as expression values of 6 genes, each with 10 measurements. I will then as if any of them have different levels of expression.

```{r make_data}
n_reps <- 10  # number of measurements per gene
expt_std_dev <- 1.5  # std. dev. of measurements
genes <- c("FAK", "talin", "paxillin", "vinculin", "B1integrin", "kindlin")
# "real" expression levels to be used as the mean in `rnorm`
real_expression_levels <- sample(seq(1, 5, 0.1), length(genes), replace = TRUE)
# create a tibble
expr_data <- tibble(gene = rep(genes, n_reps),
                    real_expr = rep(real_expression_levels, n_reps),
                    rep_num = sort(rep(1:n_reps, length(genes))))
# add in the measured expression values as a normal distribution around the mean
expr_data <- expr_data %>%
    mutate(expt_expr = rnorm(nrow(expr_data),
                             mean = real_expr,
                             sd = expt_std_dev))
head(expr_data)
```

## Statistical Tests

I then use an ANOVA to test if there are any differences between any of the comparisons of gene expression.

```{r anova}
res_aov <- aov(expt_expr ~ gene, data = expr_data)
broom::tidy(res_aov)
```

With a very low p-value from the ANOVA, I run the Tukey post-hoc test to find which genes are at different levels.

```{r tucky}
tukey_res <- TukeyHSD(res_aov)
tukey_res
```

## Plotting

Now I want to plot the `estimate` in the top-left and `p adj` in the bottom right. First, I must prepare the data for use with `geom_asymmat()` by passing the results of the Tukey post-hoc test to `asymmetrise_stats()`. You can see that it returns the data in a tibble with new columns `x` and `y` that are the result of splitting `comparison`.

```{r prep_results}
asymmat_tib <- asymmetrise_stats(tukey_res)
head(asymmat_tib)
```

Finally, I can plot the data using `geom_asymmat()`.

```{r plot_res_basic}
ggplot(asymmat_tib, aes(x = x, y = y)) +
    geom_asymmat(aes(fill_tl = estimate, fill_br = -log(adj.p.value))) +
    scale_fill_tl_gradient2(low = "dodgerblue", high = "tomato") +
    scale_fill_br_distiller(type = "seq", palette = "Greens", direction = 1)
```

And add a few styling changes with normal 'ggplot2' semantics.

```{r plot_res_full}
ggplot(asymmat_tib, aes(x = x, y = y)) +
    geom_asymmat(aes(fill_tl = estimate, fill_br = -log(adj.p.value))) +
    scale_fill_tl_gradient2(low = "dodgerblue", high = "tomato",
                            guide = guide_colourbar(order = 1)) +
    scale_fill_br_distiller(type = "seq", palette = "Greens", direction = 1,
                            guide = guide_colourbar(order = 2)) +
    theme_bw() +
    theme(panel.background = element_rect(fill = "grey50"),
          panel.grid = element_blank(),
          axis.title = element_blank(),
          plot.title = element_text(hjust = 0.5)) +
    labs(title = "Differential Gene Expression",
         fill_tl = "diff. in\nmean expr.",
         fill_br = "-log( adj. p-value )") +
    scale_x_discrete(expand = c(0,0)) +
    scale_y_discrete(expand = c(0,0))
```

One of the conclusions that can be drawn here is that the difference in expression of kindlin and FAK is the greatest and has a very low adjusted p-value. Thus, one of the conclusion is that kindlin is expressed at a lower level than FAK.

---

For more information, see the complete documentation at the ['ggasym'](https://jhrcook.github.io/ggasym/index.html) site.
